`Big data' og politologisk datavidenskab

### 2020-03-27 draft_v01.Rmd
full draft of research design section

Big data in political science research designs:
  - addressing confounding
    - it's dirty (re: Salganik) -- human behavior is mixed together with actions taken by bots/automated systems
    - drifting -- Needs of *actual* system maintainers / users may be completely orthogonal to needs of political science research (or even opposed, consider FB)
    - algorithmically confounded -- large-scale systems have robot nannies. These include everything from spell checkers to YouTube's recommendation algorithm. The observed behavior we find in big data is a consequence of the (generally) unobservable interaction between humans and these algorithms.
    - Standard sampling issues (nonrepresentative, systematic sampling bias)
  - its role
    - uncommon to directly analyze
    - more often: measurement

>>>>

Big data has applications in political science in empirical studies of all types, from description to explanation, prediction, and causal studies. Although these applications are only beginning across the social sciences, the past 15 years have provided enough experience that we can already point to one strong finding that can always guide us when we apply big data in our work: research design still matters [see @toshkov16, p. 13; @CG15]. Here we discuss aspects of research design that deserve special attention when working with big data, namely, biases in the data and the definition of the unit of analysis.

*Familiar sampling bias*

Perhaps the most heady idea about big data was expressed best in a, now infamous, article in *Wired* magazine by Chris @anderson08 -- the idea that big data eliminates sampling problems because *n*=all. That is, one can analyze *all of the data*. The article has met with a good deal of attention and pushback in the years since it was published. In early 2020, it boasts more than 2.000 citations on Google Scholar. To put that in perspective, Maurice Duverger's famous book *Political Parties*, the origin of the eponymous Duverger's Law, was published in 1959 and counts just over 7.000 citations.

As a concept, *n*=all implies that big data can be taken at face value. Its patterns reveal a complete picture of human behavior. Unfortunately, the selection problem is still with us. It arises whenever observations enter your data for reasons that systematically relate to the outcome variable. @AP08 illustrate the problem with an example from the United States' National Health Interview Survey, in which individuals are asked to rate their health on a scale. Perhaps unsurprisingly, they find that individuals in hospitals consistently rate their health as worse than those out of hospital. Then they pose an interesting question: should we conclude that hospitals make people sicker? The answer is, of course, no. We know that because we already understand that individuals in hospital *selected* into being there precisely because they were sick. Failing to recognize that would result in sampling bias.

Similar situations regularly arise with big data. For example, @harford14 recounts the story of Boston's experiment with using cell phone tracking to identify pot holes in the city's streets that need repair. The result was that the city discovered every pothole in neighborhoods frequented by young, affluent drivers--the profile of the type of person who owned a smart phone and downloaded the city's app. Uncovering a similar problem, @TSSW10 confirm a finding in previous research that Twitter mentions of German parties correlated strongly with their vote share in the 2009 parliamentary elections, with one major exception. The extremely online *Pirate Party* garned a huge number of mentions on Twitter while receiving a tiny fraction of the actual vote. In both of these cases, if *n*=all then it equals "all" of a specific, self-selected group of technology users.

Ignoring sampling bias has always meant that researchers risked finding the wrong answer. Using big data with biased samples simply means now we get extremely precise estimates of the wrong answer.

*Special headaches for big data*

Big data confronts us with new sources of bias and confounding. Perhaps the most famous example for social scientists is that of Google Flu Trends (GFT). GFT was a Google project launched in 2008 that predicted regional flu epidemics from fine-grained data on users' Google queries about likely flu symptoms [@Getal2009]. After initially receiving attention for its impressive accuracy, GFT's predictive performance began to diminish over time until--by the time the project was shuttered in 2015--for years it had produced inaccurately high forecasts5 sometimes as great as double the number of flu reported by the U.S. Centers for Disease Control and Prevention [@harford14]. The reasons for GFT's collapse are instructive for understanding both why research design still matters and what new headaches come with big data.

Several things contributed to GFT's problems, but the most noteworthy seems to have arisen after Google adjusted its main search service [see @LKKV14]. Like any organization that produces or processes big data, Google changes its service or its algorithms from time to time. In 2011 and 2012, Google added functions to its search service that suggested additional search terms to users, based on their original search terms. It even suggested possible diagnoses for search terms that involved disease symptoms. The upshot was an apparent feedback loop: users of Google search were nudged to refine their searches for symptoms and to seek out information on the flu, intensifying the signals that GFT relied on to predict flu outbreaks. There is more to GFT's story, of course, but this part of it points out some special features of big data researchers must consider to avoid sampling bias. As @salganik17 puts it, big data is *algorithmically confounded*, *dirty*, and *drifting*.

Algorithmic confounding happens when computer behavior interacts with human behavior in a system, altering the human behavior we want to study. GFT is an example, since computer behavior--the Google search algorithm's recommendations--interacted with the human behavior that GFT relied on to predict flu outbreaks (searches for flu symptoms). Google's search suggestions altered users' perceptions of their symptoms and led them to different search behavior.

By "dirty," @salganik17 means that big data also contains false positives in the form of computer behavior that is mistaken for human behavior. For example, at least a portion of Google's additional search traffic around flu syptoms may have been purely accidental, caused only by the search engine's suggestions. Smart phone users will recognize this problem as an autocorrect fail. An example that occassionaly makes headlines would be the presence of large numbers of "bots" on social media platforms. Bots are accounts that are not operated by human users, but rather are simply software programs executing automated behaviors in the form of liking, following, or commenting. Despite Twitter and other platforms' occasional purges of bot accounts, an unspecified and probably high number of bots persist. Any analysis of social media data must think carefully about how to clean the data to eliminate as much bot behavior as possible.

Drift refers to the occasional changes that occur in how big data are collected, a problem we foreshadowed in the first section. GFT was a victim of the drift in Google's search data that occurred after the introduction of search suggestions in 2011. Fundamentally, GFT's data before and after 2011 came from different populations--one in which users were treated with nudges and the other without that treatment. Therefore, these data were impossible to compare. Take note. GFT was an internal project, and yet the GFT team missed changes that their Google search colleagues introduced. Researchers using big data must stay aware of how their data may have drifted.

*Overcoming biases in big data*

These biases have no general cure, but adopting good practices can ensure you catch many problems before they become threats to the validity of your research design. First, get to know your data. If you collect your data from an API, read its documentation carefully. If you take it from a website, then understand everything you can about how the data on the site are maintained and curated--who puts it up, who can remove it, how it is updated, etc. Many organizations that produce and provide big data communicate about it, for example on company blogs or through press releases. Pay attention to these sources of information for things that can affect your research design. When in doubt, ask clarifying questions. Many system administrators or database managers are happy to help courteous researchers.

Second, try to write down your assumptions explicitly and, whenever you can, directly test them like hypotheses [see, @LBCC16]. For example, GFT could have tested for algorithmic confounding by examining whether feedback from the search engine to the user was associated with increases in symptom-related search terms. Social media analyses should test the assumption that their data are generated by humans by checking, for example, that individuals in the data do not post too repetitively, do not post inhumanly fast, and exhibit use patterns that are within some range of normal behavior for the platform.

Finally, develop a habit of running sanity checks, even when you do not anticipate problems and cannot formulate an explicit assumption. Since big data is, well, big, it is not realistic to look at the numbers or text in your data and expect to see anything useful. We do this by drawing pictures. Make histograms, scatterplots, and other plots of your variables and be sure the patterns make sense. When you discover a strange pattern, investigate it until you understand it. You may very well uncover hidden drift or the tell-tale patterns of dirty data caused by bots or other automated interference.

*Levels of analysis and measurements on big data*

Theories in empirical political science come with a *level of analysis*. For example, a theory about party behavior during election campaigns has a party-election level of analysis: each party at each election is one observation. A theory about politicians during election campaigns gives a necessarily larger number of observations per election, since each individual politician during a campaign is one observation. It is relatively rare in political science to express our theories at the level of granularity in most big data sets. It is, of course, entirely possible to develop a theory to explain individual tweets or individual laws, etc., but political science generally expresses its theories at a higher level (less granular) level of analysis.

In the next section, we introduce several popular and useful methods for working with big data. As we will see, these methods specifically associated with big data have in common that they are actually measurement tools. We will look at methods for labeling data, discovering patterns, and condensing large amounts of data into simpler, smaller data. It is typical for political scientists applying big data that their "big" data set of a million tweets or thousands of laws condenses into a data set of, for example, thousands of politician-month observations or hundreds of government-year observations.

The workhorse method of observational, quantitative hypothesis-testing in political science, even when working with big data, remains to build and estimate appropriate statistical models with appropriate control variables. Thus, we will focus on the methodological choices that distinguish analyses using big data from the more familiar social science workflow of building a purpose-built data set arranged at the correct level of analysis. Readers who are familiar with qualitative methods will be more familiar with this pattern. For example, lengthy interviews or wide-ranging case studies are often condensed into a summary coding of theoretical concepts.

This raises a point worth emphasizing about big data methods in political science. Although methods for working with big data are, necessarily, computational they do not necessarily mean that researchers must apply quantitative methodology in their overall research design.

<!-- Consider an example. Suppose you  data on the Twitter posts of Danish politicians.  -->

## measurement stuff

<!-- here we transition to the discussion of units of analysis and jusifying the focus on measurement -->

- role of big data in the research design
    - uncommon to directly analyze
    - more often: measurement

### 2020-03-27 draft_v01.tex
full draft of research design section

Big data has applications in political science in empirical studies of
all types, from description to explanation, prediction, and causal
studies. Although these applications are only beginning across the
social sciences, the past 15 years have provided enough experience that
we can already point to one strong finding that can always guide us when
we apply big data in our work: research design still matters
\citetext{\citealp[see][p.~13]{toshkov16}; \citealp{CG15}}.

@@

Perhaps the most heady idea about big data was expressed best in the,
now infamous, article in \emph{Wired} magazine by Chris
\citet{anderson08} -- the idea that big data eliminates sampling
problems because \emph{n}=all. That is, one can analyze all of the data.
The article has met with a good deal of attention and pushback in the
years since it was published. In early 2020, it boasts more than 2.000
citations on Google Scholar. To put that in perspective, Maurice
Duverger's book \emph{Political Parties}, the origin of political
science's only finding that rises to the level of a law, was published
in 1959 and counts just over 7.000 citations.

The concept of ``\emph{n}-all'' implies that big data can be taken at
face value. Its patterns reveal a complete picture of human behavior.
Unfortunately, the selection problem is still with us. It arises
whenever observations enter your data for systematic reasons.
\citet{AP08} illustrate the problem with an example from the National
Health Interview Survey in the US, in which individuals are asked to
rate their health on a scale. Perhaps surprisingly, they find that
individuals in hospitals systematically rate their health as worse than
those out of hospital. Is the obvious conclusion that hospitals make
people sicker? Of course not, individuals in hospital \emph{selected}
into being in the hospital because they were sicker.

Similar situations often arise with big data. For example,
\citet{harford14} recounts the story of Boston's experiment with using
cell phone tracking to identify pot holes in the streets that need
repair. The result was that the city discovered every pothole in
neighborhoods frequented by young, affluent drivers--they type of person
who owned a smart phone and downloaded the city's app. Likewise,
\citet{TSSW10} confirm the finding in previous research that Twitter
mentions of German parties correlated strongly with their vote share in
the 2009 parliamentary elections, with one major exception. The
extremely online \emph{Pirate Party} garned a huge number of mentions on
Twitter, while receiving a tiny fraction of the actual vote. In both of
these cases, if \emph{n}=all then it equals ``all'' of a specific,
self-selected group of technology users.

@@

Big data also confronts us with new sources of bias and confounding.
Perhaps the most famous example for social scientists is that of Google
Flu Trends (GFT). GFT was a Google project launched in 2008 that
predicted regional flu epidemics from fine-grained data on users queries
to Google about their symptoms \citep{Getal2009}. After initially
receiving attention for its impressive accuracy, GFT's predictive
performance began to diminish systematically over time until--by the
time the project was shuttered in 2015--for years it had produced
inaccurately high estimates sometimes as great as double the data
reported by US Centers for Disease Control and Prevention
\citep{harford14}. The reasons for GFT's collapse are instructive for
understanding both why research design still matters and what new
headaches come with big data.

Several things contributed to GFT's problems, but the most noteworth
seems to have arises after Google adjusted its main search service
\citep[see][]{LKKV14}. Google, like many organizations that produce and
process big data, is constantly updating its service. In 2011 and 2012,
Google added functions that suggested additional search terms to users
(based on their initial search terms) and even suggested possible
diagnoses for search terms that involved symptoms. The upshot was a
feedback loop: users of Google search were nudged to refine their
searches for symptoms and to seek out information on the flu,
intensifying the signals that GFT relied on to predict flu outbreaks.
There is more to GFT's story, of course, but this highlights some
important principles for researchers to keep in mind.

Big data in political science research designs: - addressing confounding
- it's dirty (re: Salganik) -- human behavior is mixed together with
actions taken by bots/automated systems - drifting -- Needs of
\emph{actual} system maintainers / users may be completely orthogonal to
needs of political science research (or even opposed, consider FB) -
algorithmically confounded -- large-scale systems have robot nannies.
These include everything from spell checkers to YouTube's recommendation
algorithm. The observed behavior we find in big data is a consequence of
the (generally) unobservable interaction between humans and these
algorithms.

>>>>

Big data has applications in political science in empirical studies of
all types, from description to explanation, prediction, and causal
studies. Although these applications are only beginning across the
social sciences, the past 15 years have provided enough experience that
we can already point to one strong finding that can always guide us when
we apply big data in our work: research design still matters
\citetext{\citealp[see][p.~13]{toshkov16}; \citealp{CG15}}. Here we
discuss aspects of research design that deserve special attention when
working with big data, namely, biases in the data and the definition of
the unit of analysis.

@@

Perhaps the most heady idea about big data was expressed best in a, now
infamous, article in \emph{Wired} magazine by Chris \citet{anderson08}
-- the idea that big data eliminates sampling problems because
\emph{n}=all. That is, one can analyze \emph{all of the data}. The
article has met with a good deal of attention and pushback in the years
since it was published. In early 2020, it boasts more than 2.000
citations on Google Scholar. To put that in perspective, Maurice
Duverger's famous book \emph{Political Parties}, the origin of the
eponymous Duverger's Law, was published in 1959 and counts just over
7.000 citations.

As a concept, \emph{n}=all implies that big data can be taken at face
value. Its patterns reveal a complete picture of human behavior.
Unfortunately, the selection problem is still with us. It arises
whenever observations enter your data for reasons that systematically
relate to the outcome variable. \citet{AP08} illustrate the problem with
an example from the United States' National Health Interview Survey, in
which individuals are asked to rate their health on a scale. Perhaps
unsurprisingly, they find that individuals in hospitals consistently
rate their health as worse than those out of hospital. Then they pose an
interesting question: should we conclude that hospitals make people
sicker? The answer is, of course, no. We know that because we already
understand that individuals in hospital \emph{selected} into being there
precisely because they were sick. Failing to recognize that would result
in sampling bias.

Similar situations regularly arise with big data. For example,
\citet{harford14} recounts the story of Boston's experiment with using
cell phone tracking to identify pot holes in the city's streets that
need repair. The result was that the city discovered every pothole in
neighborhoods frequented by young, affluent drivers--the profile of the
type of person who owned a smart phone and downloaded the city's app.
Uncovering a similar problem, \citet{TSSW10} confirm a finding in
previous research that Twitter mentions of German parties correlated
strongly with their vote share in the 2009 parliamentary elections, with
one major exception. The extremely online \emph{Pirate Party} garned a
huge number of mentions on Twitter while receiving a tiny fraction of
the actual vote. In both of these cases, if \emph{n}=all then it equals
``all'' of a specific, self-selected group of technology users.

@@

Big data confronts us with new sources of bias and confounding. Perhaps
the most famous example for social scientists is that of Google Flu
Trends (GFT). GFT was a Google project launched in 2008 that predicted
regional flu epidemics from fine-grained data on users' Google queries
about likely flu symptoms \citep{Getal2009}. After initially receiving
attention for its impressive accuracy, GFT's predictive performance
began to diminish over time until--by the time the project was shuttered
in 2015--for years it had produced inaccurately high forecasts5
sometimes as great as double the number of flu reported by the U.S.
Centers for Disease Control and Prevention \citep{harford14}. The
reasons for GFT's collapse are instructive for understanding both why
research design still matters and what new headaches come with big data.

Several things contributed to GFT's problems, but the most noteworthy
seems to have arisen after Google adjusted its main search service
\citep[see][]{LKKV14}. Like any organization that produces or processes
big data, Google changes its service or its algorithms from time to
time. In 2011 and 2012, Google added functions to its search service
that suggested additional search terms to users, based on their original
search terms. It even suggested possible diagnoses for search terms that
involved disease symptoms. The upshot was an apparent feedback loop:
users of Google search were nudged to refine their searches for symptoms
and to seek out information on the flu, intensifying the signals that
GFT relied on to predict flu outbreaks. There is more to GFT's story, of
course, but this part of it points out some special features of big data
researchers must consider to avoid sampling bias. As \citet{salganik17}
puts it, big data is \emph{algorithmically confounded}, \emph{dirty},
and \emph{drifting}.

Algorithmic confounding happens when computer behavior interacts with
human behavior in a system, altering the human behavior we want to
study. GFT is an example, since computer behavior--the Google search
algorithm's recommendations--interacted with the human behavior that GFT
relied on to predict flu outbreaks (searches for flu symptoms). Google's
search suggestions altered users' perceptions of their symptoms and led
them to different search behavior.

By ``dirty,'' \citet{salganik17} means that big data also contains false
positives in the form of computer behavior that is mistaken for human
behavior. For example, at least a portion of Google's additional search
traffic around flu syptoms may have been purely accidental, caused only
by the search engine's suggestions. Smart phone users will recognize
this problem as an autocorrect fail. An example that occassionaly makes
headlines would be the presence of large numbers of ``bots'' on social
media platforms. Bots are accounts that are not operated by human users,
but rather are simply software programs executing automated behaviors in
the form of liking, following, or commenting. Despite Twitter and other
platforms' occasional purges of bot accounts, an unspecified and
probably high number of bots persist. Any analysis of social media data
must think carefully about how to clean the data to eliminate as much
bot behavior as possible.

Drift refers to the occasional changes that occur in how big data are
collected, a problem we foreshadowed in the first section. GFT was a
victim of the drift in Google's search data that occurred after the
introduction of search suggestions in 2011. Fundamentally, GFT's data
before and after 2011 came from different populations--one in which
users were treated with nudges and the other without that treatment.
Therefore, these data were impossible to compare. Take note. GFT was an
internal project, and yet the GFT team missed changes that their Google
search colleagues introduced. Researchers using big data must stay aware
of how their data may have drifted.

\emph{Overcoming biases in big data}

These biases have no general cure, but adopting good practices can
ensure you catch many problems before they become threats to the
validity of your research design. First, get to know your data. If you
collect your data from an API, read its documentation carefully. If you
take it from a website, then understand everything you can about how the
data on the site are maintained and curated--who puts it up, who can
remove it, how it is updated, etc. Many organizations that produce and
provide big data communicate about it, for example on company blogs or
through press releases. Pay attention to these sources of information
for things that can affect your research design. When in doubt, ask
clarifying questions. Many system administrators or database managers
are happy to help courteous researchers.

Second, try to write down your assumptions explicitly and, whenever you
can, directly test them like hypotheses \citep[see,][]{LBCC16}. For
example, GFT could have tested for algorithmic confounding by examining
whether feedback from the search engine to the user was associated with
increases in symptom-related search terms. Social media analyses should
test the assumption that their data are generated by humans by checking,
for example, that individuals in the data do not post too repetitively,
do not post inhumanly fast, and exhibit use patterns that are within
some range of normal behavior for the platform.

Finally, develop a habit of running sanity checks, even when you do not
anticipate problems and cannot formulate an explicit assumption. Since
big data is, well, big, it is not realistic to look at the numbers or
text in your data and expect to see anything useful. We do this by
drawing pictures. Make histograms, scatterplots, and other plots of your
variables and be sure the patterns make sense. When you discover a
strange pattern, investigate it until you understand it. You may very
well uncover hidden drift or the tell-tale patterns of dirty data caused
by bots or other automated interference.

\emph{Levels of analysis and measurement}
